Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-24 resubmission/02-antecedents.qmd
Seccion caso Chileno

>>>>

## The case of Chile

Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@huber_democracy_2012]. Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@ferre_welfare_2023]. This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives. Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@torche_unequal_2005]. While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@EspinozaNunez2014].
Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2019]. In this context, Chile experienced a wave of mass protests from October 2019 to March 2020. Initially led by high school students, the demonstrations quickly gained momentum as broader segments of society mobilized around demands for more equality of opportunities, particularly regarding precarious access to welfare services. The political system interpreted this unrest as a call to reform existing political institutions, with a special focus on equalizing opportunities and providing more egalitarian access to welfare by reducing reliance on out-of-pocket mechanisms [@somma_no_2021].

Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023. At the same time, around 55% and 47% of the public between 2016 and 2023 disagreed that effort and talent are rewarded accordingly. This coincides with 90% of citizens agreeing that income differences in Chile are too large [@castillo_changes_2025].
In sum, Chile represents an illustrative case of a particular form of moral economy [@koos_moral_2019], in which socioeconomic inequalities, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.
